---
title: "Exploring different prior choices for the expectation-maximization algorithm."
author: "Damon Pham, Kushal Shah, Ethan Violette"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
list.of.packages <- c('kableExtra', 'knitr', 'ggplot2','gridExtra','reshape2','grid')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)
rm(list.of.packages, new.packages)

opts_chunk$set(echo=FALSE, cache=TRUE, autodep=TRUE,
               message=FALSE, warning=FALSE,
               fig.align="center")

source('scripts.R')
```

Our own implementation converges correctly when the initial prior distributions are within range of the true prior distributions. Below are plots showing the data distribution (x) and a sample from the solved MOG (y). Clearly, the distributions are near-identical.

```{r}
set.seed(0)
x = generate.mog(n.data=1000, mu=c(0,.8, .3), sigma=c(.01,.01,.02), prior=c(.6,.2,.2))
hist(x, breaks=100)

ans1 = EM(data=x, n.classes=3, prior.choice='equal.priors.random.params', max.iter=1000)

y = generate.mog(n.data=1000, mu=as.numeric(ans1$mu), sigma=ans1$sigma, prior=ans1$prior)
hist(y, breaks=100)
```

However, when the true prior distributions are on a different magnitude, the algorithm fails to converge.

```{r}
set.seed(0)
x = generate.mog(n.data=1000, mu=c(0,8, 3), sigma=c(.01,.01,.02), prior=c(.6,.2,.2))
hist(x, breaks=100)

ans1 = EM(data=x, n.classes=3, prior.choice='equal.priors.random.params', max.iter=1000)

y = generate.mog(n.data=1000, mu=as.numeric(ans1$mu), sigma=ans1$sigma, prior=ans1$prior)
hist(y, breaks=100)
```

